Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-05-13 paper.tex

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), 
PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), RPPL~\cite{1174106}, SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters\closednote{Tristan}{Reverted from Marc's 'cluster
  scheduler'}{}{}, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance
tracking, \closednote{Naj}{should we not mention PMP which is used for
  CIVET?}{Tristan}{Done, as discussed.}  redundancy detection across
analyses to avoid re-computation, and so on. A wide array of workflows
have been implemented using these pipeline systems and are now shared across
neuroimaging groups world-wide, which represents a tremendous
opportunity for science gateways to leverage. Domain-specific engines
nicely complement e-Science systems that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipeline systems.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways are tailored to a particular
engine while other ones are more general and host applications
executed by different types of engines. 
%Some gateways may not even use any
%workflow engine and instead focus on the execution of simple tasks.

@@

Scalability is also a crucial concern for such multi-user,
high-throughput systems. For this purpose, science gateways may
balance the load among different instances of the same engine, start
new engines elastically, and use advanced task scheduling policies on
the infrastructure to improve performance, fault-tolerance and
fairness among users.

@@

Other specific features may also be available, for instance
data vizualisation and quality control, workflow edition, debugging
instruments, social tools among users, etc.

@@

\paragraph{Scalability} The service architecture supports multiple
engine instances through multiple workflow services. In VIP for
instance, this feature has been available from release 1.17. A basic
load-balancing mechanism is available that sends new workflow
executions to the engine instance that has the least active
executions. To avoid ``black-hole'' syndromes created by failing
engine instances, engine instances are automatically disabled when
workflows cannot be submitted to them. Adding a new engine instance,
however, requires manual intervention to declare the new instance in
the science gateway (\texttt{S$_1$}=1). Consequently, elastic engines
are difficult to implement because they require a mechanism to update
the science gateway configuration when a new engine instance is
available. Distributing the execution of a single workflow in multiple
engines is usually not possible unless the workflow engine has
specific abilities (\texttt{S$_2$}=1). The scheduling of tasks on the
infrastructure is as complex as in any other architecture since the
workflow engine might implement any kind of scheduling policy
(\texttt{S$_3$}=0).

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). This is a
major interest of task encapsulation. Distributed engines are
not supported by default (\texttt{S$_2$}=1). Task scheduling is
slightly more complex than in the other approaches due to the special
role of the task that executes the workflow engine
(\texttt{S$_3$}=1). Indeed, the reliability of this task is critical
since all the sub-tasks in the workflow depend on it and, depending on
the recovery capabilities of the workflow engine, may need to be
resubmitted if the workflow task fails. The workflow task is also
longer than all its sub-tasks, which increases its chances of
failure. In addition, task parameters, for instance estimated
walltime, are more difficult to estimate for the workflow task than
for the sub-tasks, which may generate issues such as selection of wrong
batch queues on clusters. Finally, the interdependencies between the
workflow task and its sub-tasks may create deadlocks when there is
contention. For instance, if only 1 computing resource is
available for the science gateway and if the workflow task is
running on it and submits sub-tasks, then the sub-tasks could only
execute when the resource is available, which will never happen
because the workflow task will not complete until the sub-tasks
complete. This configuration can be generalized to an infrastructure
with $n$ resources where $n$ workflows are submitted. In practice,
however, the number of submitted workflows usually remains lower than
the number of computing resources available on this infrastructure,
which makes such deadlocks unlikely to happen.

@@

\paragraph{Scalability} New engine instances only require new agents,
which is easily automated (\texttt{S$_1$=0}) and by design very
suitable for elastic computing. For instance, auto-scaling rules can
be implemented to start new agents when the workload in the science
gateway exceeds a certain threshold~\cite{lorido2012auto}. Distributed
engines are not available by default (\texttt{S$_2$=1}) and task
scheduling is as complex as in any other architecture
(\texttt{S$_3$=0}).

@@

Regarding extensibility, most architectures are overall comparable,
except nested workflows (with service invocation) which is
significantly behind. This is explained by the complexity of the
nested workflow architecture, with 2 infrastructures and 2 workflow
services. Task encapsulation and workflow import (with service invocation)
perform a bit better when integrating new engines (\texttt{E$_1$}),
which is interesting given the current profusion of engines with
different characteristics.  However, task encapsulation, workflow import
(with service invocation) and tight integration perform worse than the
others when adding engine versions (\texttt{E$_2$}) due to the need to
update infrastructure (task encapsulation), science gateway (tight
integration), or workflow converter (workflow import). Workflow import
performs worse than the others when integrating new workflows
(\texttt{E$_3$}) because of the language conversion step. All
architectures except nested workflows are equivalent when
integrating new infrastructures (\texttt{E$_4$}).

The pool architecture is overall the most scalable, which is not a
surprise since it was designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same overall. The global scalability score, however, should not
mask the unique characteristics of architectures regarding this
criterion. Nested workflows are the only architecture that can easily
accommodate distributed workflow execution, which can be critical in
some cases. At the same time, the scheduling constraints created by
task encapsulation and nested workflows could well become
showstoppers depending on the type of targeted
infrastructure. Non-reliable infrastructures, for example, could
hardly cope with workflow engines being wrapped in computing tasks as
done in task encapsulation. The availability of multiple engine instances
could also become a critical feature for science gateways with
important workloads, which would favor task encapsulation and pool.

Differences in other specific features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows,
which may be interesting in some cases. Also, task encapsulation and nested
workflows may complexify fine-grained debugging. \closednote{Rafael}{I would
  suggest to provide some more examples when this could be
  interesting.}{Tristan}{Done.}  The availability of fine-grained debugging
information may also be critical to efficient user support.

@@

A few limitations should be considered when using the
results of our evaluation. First of all, the evaluation is done at a
quite high-level of abstraction that may be a bit artificial when
applied to real systems. Abstracting architectures from their
implementation in specific systems is useful to compare them, but we
also realize that it comes at the cost of realism. In particular, the
presented architectures are theoretical patterns that may be blended
in actual systems. The distinction between tight integration and
service invocation, for instance, may not always be that clear in
practice. Service invocation may also be combined with task
encapsulation in some cases. However, the criteria discussed in this
work would still apply to broadly compare and categorize such cases of
hydrid architectures.

Moreover, all the interactions involved in the architectures were
treated equally while some of them are obviously more complex
than others. For example, interaction \texttt{g} (workflow language
conversion) is clearly more complex than interaction \texttt{d}
(service invocation), and may create much more reliability issues as
well. However, quantifying the robustness and amount of development
associated with each interaction and software component cannot
realistically be done without entering into the details of a
particular system.

The particular case of interaction \texttt{g} (workflow language
conversion) should be further discussed since it is very likely that
this interaction could not be easily generalized to any workflow
language. For instance, converting FSL, PSOM or Nipype pipelines to
any other workflow language is problematic because these engines rely
on general-purpose programming languages such as Bash, Octave and
Python, which are much richer than scientific workflow languages.

The abstract nested workflows and workflow import patterns were
instantiated with service invocation so that they can be analyzed in
the same framework as the other architectures. Other types of
instantiation, for instance nested workflows with task encapsulation,
could also be envisaged. We chose to limit ourselves to instantiations
with service invocation because the resulting architectures are
implemented in real systems, and because service invocation is largely
used. Nevertheless, it could be interesting to explore other types of
instantiations, in particular nested workflows with task
encapsulation.

The particular technical or historical context of a science gateway
project may obviously influence the choice of an architecture to
integrate workflow engines. For instance, some workflow engines may be
already available as web services, which would tend to favor service
invocation, and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \texttt{c}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.

@@

% Summary of contributions
We reviewed architectures used to integrate workflow engines in
science gateways, we described them in a system-independent framework
suitable for comparison, we illustrated them on real systems, we
developed quantitative metrics for evaluation, and we discussed the
pros and cons of all the presented architectures based on these metrics.
\closedanswerednote{Rafael}{One thing that is missing in the
  conclusion or the limitations section is to state that this work
  does not intend to quantify the amount of work required to develop
  science gateways using a specific architecture, but the goal is to
  highlight what are the main components and interactions for each
  architecture, and the main advantages and drawbacks. This should
  also be stated in the introduction.}{Tristan}{This is in
  contradiction with the metric on 'integration effort'. What do you
  think?}{Tristan}{Fixed, as discussed in github \#8.}

% Impact of the work
To the best of our knowledge, our work is the first of its kind. So
far, the literature on science gateways and workflow engines has
focused on the description of particular systems, or on the
presentation of a particular architecture.  Instead, our analysis
abstracts and evaluates architectural patterns independently from any
particular system. It provides general insight to integrate science
gateways and workflow engines, which we hope will be useful for
software architects.

>>>>

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), 
PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), RPPL~\cite{1174106}, SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters\closednote{Tristan}{Reverted from Marc's 'cluster
  scheduler'}{}{}, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance
tracking, \closednote{Naj}{should we not mention PMP which is used for
  CIVET?}{Tristan}{Done, as discussed.} or  redundancy detection across
analyses to avoid re-computation. A wide array of workflows
have been implemented using these pipeline systems and are now shared across
neuroimaging groups world-wide, which represents a tremendous
opportunity for science gateways to leverage. Domain-specific engines
nicely complement e-Science systems that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}), \closednote{Naj}{I think and is redundant here, you can replace is with a comma}{Tristan}{yes!} using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipeline systems.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways are tailored to a particular
engine, while other ones are more general and host applications
executed by different types of engines. %Some gateways may not even use any
%workflow engine and focus on the execution of simple tasks.

@@

Scalability is also a crucial concern for such multi-user,
high-throughput systems. For this purpose, science gateways may
balance the load among different instances of the same engine, start
new engines elastically using auto-scaling techniques such as the ones
reviewed in~\cite{lorido2012auto}, and use advanced task scheduling
policies on the infrastructure to improve performance, fault-tolerance
and fairness among users.

@@

Other specific features may also be available, for instance
data vizualisation and quality control, workflow edition, debugging
instruments, or social tools among users. \closednote{Naj}{I think using etc after such as, for example and e.g. is not correct. --sorry for being pedantic}{Tristan}{You are completely right, this is now fixed.}

@@

\paragraph{Scalability} The service architecture supports multiple
engine instances through multiple workflow services. In VIP for
instance, this feature has been available from release 1.17. A basic
load-balancing mechanism is available that sends new workflow
executions to the engine instance which has the least active
executions. To avoid ``black-hole'' syndromes created by failing
engine instances, engine instances are automatically disabled when
workflows cannot be submitted to them. Adding a new engine instance,
however, requires manual intervention to declare the new instance in
the science gateway (\texttt{S$_1$}=1). Consequently, elastic engines
are difficult to implement because they require a mechanism to update
the science gateway configuration when a new engine instance is
available. Distributing the execution of a single workflow in multiple
engines is usually not possible unless the workflow engine has
specific abilities (\texttt{S$_2$}=1). The scheduling of tasks on the
infrastructure is as complex as in any other architecture since the
workflow engine might implement any kind of scheduling policy
(\texttt{S$_3$}=0).

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). This is a
major interest of task encapsulation. Distributed engines are not
supported by default (\texttt{S$_2$}=1). Task scheduling is slightly
more complex than in the other approaches due to the special role of
the task that executes the workflow engine (\texttt{S$_3$}=1). Indeed,
the reliability of this task is critical since all the sub-tasks in
the workflow depend on it and, depending on the recovery capabilities
of the workflow engine, may need to be resubmitted if the workflow
task fails. The workflow task is also longer than all its sub-tasks,
which increases its chances of failure. In addition, task parameters,
for instance estimated walltime, are more difficult to estimate for
the workflow task than for the sub-tasks because workflows are by
definition more complex than their sub-tasks: errors on sub-task
parameter estimations accumulate in the workflow, and additional
control constructs such as tests and loops may further increase the
uncertainty. Such parameter estimation errors may generate issues such
as selection of wrong batch queues on clusters or task termination due
to exceeded quotas. \closednote{Naj}{The lay reader wonders why?  an
  example like the one offered for interdependencies can
  help}{Tristan}{Done, thanks.} Finally, the interdependencies between
the workflow task and its sub-tasks may create deadlocks when there is
contention. For instance, if only 1 computing resource is available
for the science gateway and if the workflow task is running on it and
submits sub-tasks, then the sub-tasks could only execute when the
resource is available, which will never happen because the workflow
task will not complete until the sub-tasks complete. This
configuration can be generalized to an infrastructure with $n$
resources where $n$ workflows are submitted. In practice, however, the
number of submitted workflows usually remains lower than the number of
computing resources available on this infrastructure, which makes such
deadlocks unlikely to happen. \closednote{Naj}{so is this the solution to
  fixing the above mentioned limitations with
  scalability?}{Tristan}{Unfortunately this is not a solution. We just
  say that this situation is unlikely to happen which is true.}

@@

\paragraph{Scalability} New engine instances only require new agents,
which is easily automated (\texttt{S$_1$=0}) and by design very
suitable for elastic computing.\closednote{Naj}{I feel the concept of elastic computing should be mentioned in introductory sections}{Tristan}{You are right, this is now fixed.} For instance, auto-scaling rules can
be implemented to start new agents when the workload in the science
gateway exceeds a certain threshold~\cite{lorido2012auto}. Distributed
engines are not available by default (\texttt{S$_2$=1}) and task
scheduling is as complex as in any other architecture
(\texttt{S$_3$=0}).

@@

Regarding extensibility, most architectures are overall comparable,
except nested workflows (with service invocation) which are
significantly behind. This is explained by the complexity of the
nested workflow architecture, with 2 infrastructures and 2 workflow
services. Task encapsulation and workflow import (with service invocation)
perform slightly better when integrating new engines (\texttt{E$_1$}),
which is interesting given the current profusion of engines with
different characteristics.\note {Naj}{what is interesting?}  However, task encapsulation, workflow import
(with service invocation) and tight integration perform worse than the
others when adding engine versions (\texttt{E$_2$}), due to the need to
update infrastructure (task encapsulation), science gateway (tight
integration), or workflow converter (workflow import). Workflow import
performs worse than the others when integrating new workflows
(\texttt{E$_3$}) because of the language conversion step. All
architectures except nested workflows are equivalent when
integrating new infrastructures (\texttt{E$_4$}).

The pool architecture is overall the most scalable, which is not surprising since it is designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same overall. The global scalability score, however, should not
mask \note{Naj}{what do you mean by mask?}the unique characteristics of architectures regarding this
criterion. Nested workflows are the only architecture that can easily
accommodate distributed workflow execution, which can be critical in
some cases. At the same time, the scheduling constraints created by
task encapsulation and nested workflows could well become
showstoppers \note{Naj}{please don't use these from-film-marketing words?! :-) }depending on the type of targeted
infrastructure. Non-reliable infrastructures \note{Naj}{you mean by the showstoppers, i.e. the nested workflows which can accommodate distributed executions but are limited by scheduling constraints?} , for example, could
hardly cope with workflow engines being wrapped in computing tasks as
done in task encapsulation. The availability of multiple engine instances
could also become a critical feature for science gateways with
important workloads, which would favor task encapsulation and pool. \note{Naj}{I feel this paragraph is a bit vague, and doesn't distill and contextualize the thorough work described before.}

Differences in other specific features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows,
which may be interesting in some cases \note{Naj}{Which?}. Also, task encapsulation and nested
workflows may complexify fine-grained debugging. \closednote{Rafael}{I would
  suggest to provide some more examples when this could be
  interesting.}{Tristan}{Done.}  The availability of fine-grained debugging
information may also be critical to efficient user support.

@@

A few limitations should be considered when using the results of our
evaluation. First, our evaluation methods aimed to provide comparative
metrics, however the high-level of abstraction used to derive these
metrics hinders the complexity of real systems to some
extent.\closednote{Naj}{Here, you were bashing your article harshly,
  so I rewrote it}{Tristan}{yup, thanks!}. In particular, the
presented architectures are abstract patterns that may be mixed
together in actual systems.\answerednote{Naj}{something about "theoretical
  pattern", and "blending" is jarring, are these common words in the
  computer sci context?}{Tristan}{Better now?} The distinction between tight integration and
service invocation, for instance, may not always be that clear in
practice. Service invocation may also be combined with task
encapsulation in some cases. However, the criteria discussed in this
work would still apply to broadly compare and categorize such cases of
hybrid architectures.

Moreover, all the interactions involved in the architectures were
treated equally while some of them are obviously more complex than
others. For example, interaction \texttt{g} (workflow language
conversion) is clearly more complex than interaction \texttt{d}
(service invocation), and may create much more reliability issues as
well. However, without entering into the details of a particular
system, quantification of the robustness or the amount of development
associated with each interaction and software component will not be
precise.

The particular case of interaction \texttt{g} (workflow language
conversion) should be further discussed \note{Naj}{so are you discussing it here, or in the future?}since 
this interaction may not be easily generalized to any workflow
language. For instance, converting FSL, PSOM or Nipype pipelines to
any other workflow language is problematic because these engines rely
on general-purpose programming languages such as Bash, Octave and
Python, which are much richer than scientific workflow languages.

The abstract nested workflows and workflow import patterns were
instantiated with service invocation so that they can be analyzed in
the same framework as the other architectures. Other types of
instantiation, for instance nested workflows with task encapsulation,
could also be envisaged. We chose to limit ourselves to instantiations
with service invocation because the resulting architectures are
implemented in real systems, and because service invocation is largely
used. Nevertheless, it could be interesting to explore other types of
instantiations, in particular nested workflows with task
encapsulation.\note{Naj}{I think in the discussion, you may want to close the circle on a statement in introduction, which referred to your 7-year experience with CBRAIN and VIP}

The particular technical or historical context of a science gateway
project may obviously influence the choice of an architecture to
integrate workflow engines. For instance, some workflow engines may be
already available as web services\note{Naj}{This is a good place to offer some examples}, which would tend to favor service
invocation, and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \texttt{c}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.

@@

% Summary of contributions
We have systematically reviewed architectures used to integrate workflow engines in
science gateways. These architectures are described in a system-independent framework
suitable for comparison, illustrated on real systems, and evaluated using novel quantitative metrics that allow simple comparison across architectures. We have discussed the
pros and cons of all the presented architectures based on these metrics.
\closedanswerednote{Rafael}{One thing that is missing in the
  conclusion or the limitations section is to state that this work
  does not intend to quantify the amount of work required to develop
  science gateways using a specific architecture, but the goal is to
  highlight what are the main components and interactions for each
  architecture, and the main advantages and drawbacks. This should
  also be stated in the introduction.}{Tristan}{This is in
  contradiction with the metric on 'integration effort'. What do you
  think?}{Tristan}{Fixed, as discussed in github \#8.}

% Impact of the work
To the best of our knowledge, our work is the first of its kind.\note{Naj}{instead of its kind, explain what kind it is, e.g. systematic review of all ...} So
far, the literature on science gateways and workflow engines has
focused on the description of particular systems, or on the
presentation of a particular architecture.  Instead, our analysis
abstracts and evaluates architectural patterns independently from any
particular system. It provides general insight to integrate science
gateways and workflow engines, which we hope will be useful for
software architects.
